---
title: "Marine Stranding Analysis -- Britney's File #1 :) -- Odontocetes"
author: "Britney Pepper"
date: "4/2/2022"
output: html_document
  html_document:
    df_print: paged
---

#Setup
```{r Setup, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Working Directory
getwd()

#Load packages
library(tidyverse)
library(tidyr)
library(lubridate)
library(dplyr)
library(cowplot)
library(ggplot2)
library(ggridges)
library(agricolae)
library(lubridate)
library(corrplot)
library(htmltools)
library(sf)
library(leaflet)
library(mapview)
library(RColorBrewer)
library(leafsync)
library(trend)
library(zoo)
library(Kendall)
library(tseries)
library(knitr)

#Set theme
mytheme <- theme_bw() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(mytheme)
```


#Load Data
```{r Data}
stranding_data <- read.csv("../Data/Processed/strandings_final.csv", stringsAsFactors = TRUE)

#Odontocetes Data
odontocete_stranding_data <- stranding_data %>%
  filter(Family == "Odontocetes") %>%
  select(Family:longitude)

#Setting the date
odontocete_stranding_data$Date <- as.Date(odontocete_stranding_data$Date, format = "%Y-%m-%d")


#Annual Strandings
annual_strandings_totals <- read.csv("../Data/Processed/annual_strandings_totals.csv", stringsAsFactors = TRUE)

annual_odontocetes_strandings <- annual_strandings_totals %>%
  select(Year, Odontocetes_Total)

#Monthly Strandings
monthly_strandings_totals <- read.csv("../Data/Processed/monthly_strandings_totals.csv", stringsAsFactors = TRUE)

monthly_odontocetes_strandings <- monthly_strandings_totals %>%
  select(Month, Odontocetes_Total)


#Odonotcete stranding numbers
odontocete_stranding_numbers <- odontocete_stranding_data %>%
                    filter(Family == "Odontocetes") %>%
                    group_by(Month, Year) %>%
                    summarise(Total = n(), .groups='drop') %>%
                    mutate(Date = my(paste0(Month,"-",Year))) %>%
                    select(Date, Total)

Date_Range <- as.data.frame(seq.Date(from = as.Date("1990-01-01"), to = as.Date("2011-12-01"), by = "months"))
names(Date_Range)[1] <- "Date"

odontocete_stranding_numbers_final <- left_join(Date_Range, odontocete_stranding_numbers, by = "Date" )

odontocete_stranding_numbers_final[is.na(odontocete_stranding_numbers_final)] = 0

```



#Odontocete Data Exploration
```{r Odontocete Data Exploration}
summary(odontocete_stranding_data)

#By year
ggplot(odontocete_stranding_data) +
  geom_bar(aes(x = Year, fill = common_name)) +
  ylab("Count") +
  labs(title = "Total Odontocete Strandings per Year by Species") +
  scale_fill_discrete(name = "Odontocete Species", labels = c("Atlantic White-sided Dolphin", "Blainville's Beaked Whale","Common Bottlenose Dolphin", "Dwarf Sperm Whale", "Harbor Porpoise","Long-finned Pilot Whale", "Pygmy Sperm Whale", "Risso's Dolphin", "Short-beaked Common Dolphin", "Short-finned Pilot Whale", "Sowerby's Beaked Whale", "Striped Dolphin", "Unidentified Toothed Whales"))

#By month
ggplot(odontocete_stranding_data) +
  geom_bar(aes(x = Month, fill = common_name)) +
  ylab("Count") +
  labs(title = "Total Odontocete Strandings per Month by Species") + 
  scale_fill_discrete(name = "Odontocete Species", labels = c("Atlantic White-sided Dolphin", "Blainville's Beaked Whale","Common Bottlenose Dolphin", "Dwarf Sperm Whale", "Harbor Porpoise","Long-finned Pilot Whale", "Pygmy Sperm Whale", "Risso's Dolphin", "Short-beaked Common Dolphin", "Short-finned Pilot Whale", "Sowerby's Beaked Whale", "Striped Dolphin", "Unidentified Toothed Whales")) +
  scale_x_discrete(name = "Month",
                   limits = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"))
```

There appears to be an increasing trend in the total number of odontocete strandings as the years progress. The odontocete strandings have the lowest amount in 1993 and the highest in 2011, where the count gradually increases over the years. The different species that are stranded throughout the years do not appear to have a noticeable trend. When looking at the strandings per month, the months with the most strandings are during March, May, and July. There may be possible seasonal trends, where the spring and early summer see a greater number of strandings than the other months due to calving season.



```{r, message = FALSE}
odontocete_stranding_plot <-
ggplot(annual_odontocetes_strandings, aes(x = Year, y = Odontocetes_Total)) +
  geom_line(color = "red") +
  ylab("Pinniped Strandings") +
  geom_smooth(method = lm) +
  labs(title = "Odontocete Strandings Per Year")

print(odontocete_stranding_plot)
```
When looking at the graph above, you can see the trend in the number of strandings of odontocetes over the years, and how they are increasing. All of these strandings can be seen in the map below.



```{r Geospatial Exploration}
odontocete_sf <- odontocete_stranding_data %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

mapview(odontocete_sf, layer.name = "Odontocete Strandings")
```



#Odontocete Analysis
##Statisical Analysis

Null Hypothesis 1: There is no effect of individual year on the total number of odontocete strandings.
Alternative Hypothesis 1: There is an effect of year on the total number of odontocete strandings.

Null Hypothesis 2: There is no effect of month on the total number of odontocete strandings.
Alternative Hypothesis 2: There is an effect of month on the total number of odontocete strandings.

```{r GLM Models}

odontocete_model <- glm(data = annual_odontocetes_strandings, Odontocetes_Total ~ factor(Year), family = poisson)
summary(odontocete_model)

odontocete_model2 <- glm(data = monthly_odontocetes_strandings, Odontocetes_Total ~ factor(Month), family = poisson)
summary(odontocete_model2)
```

We reject the null hypothesis that there is no effect of year on the total number of odontocete strandings. Of all the years in the odontocete strandings, only 2 years are statistically significant; 1990 and 2011 were both statistically significant with p-values of 1.14e-05 and 0.0198 respectively. The null deviance was had a value of 4.0129e+01 and the residual deviance had a value of. -2.1867e-21.

We reject the null hypothesis that there is no effect of month on the total number of odontocete strandings. There were four months in which the odonotocete strandings were statistically significant: January, March, May, and June.

This indicates that there is a difference in the number of odontocete strandings across the years and months of the study. This could be explained by increase boat and shipping traffic as the years progress that contribute to increased ocean noise and interference with their communication that have led them to become stranded. Seasonality may not have an effect because of the consistency of boat and ship traffic throughout each year.




We reject the null hypothesis that individual year has no effect on sea turtle strandings. The sea turtle strandings in 9 of the 21 years are not significantly different (p-values: 0.147 - 0.706). 1995 was the most significantly different (p-value: 0.000236). 1991 (p-value: 0.0056), 1993 (p-value: 0.0078), 2008 (p-value: 0.0022), 2010 (p-value: 0.0016), and 2011 (p-value: 0.003) were the next most statitistically different years. The null deviance of this model is 7.4606e+01, which indicates ??? The residual deviance of this model is -4.8850e-15, which indicates ??? 

We reject the null hypothesis that month has no effect on sea turtle strandings. Of the 12 months, 5 of months were significantly different. May (p-value: 0.000269), June (p-value: 3.98e-06), July (p-value: 1.91e-05), August (p-value: 0.004106), and September (p-value: 0.049935) were significantly different. This indicates that sea turtle strandings are concentrated to the summer months in our study area. The null deviance of this model is 4.9235e+02, which indicates ??? The residual deviance of this model is 3.0332e-10, which indicates ???


#Geospatial Analysis
```{r Geospatial Analysis}
#Transform the data to UTM 19
odontocete_sf_utm <- st_transform(odontocete_sf,32619)
#The data points straddle UTM zones 18 and 19, zone 19 was chosen because that is where the greatest concentration of points seems to be located.

odontocete_sf_utm$geometry


odo_long_mean <- mean(odontocete_stranding_data$longitude)
odo_lat_mean <- mean(odontocete_stranding_data$latitude)

odontocete_stranding_mean_point <- data.frame(odo_long_mean, odo_lat_mean)

odontocete_mean_sf <- odontocete_stranding_mean_point %>%
  st_as_sf(coords = c("odo_long_mean", "odo_lat_mean"), crs = 4326)

odontocete_mean_utm <- st_transform(odontocete_mean_sf,32619)

odontocete_distance_matrix <- st_distance(odontocete_sf_utm, odontocete_sf_utm, pairwise = T)

odo_mean_distance <-  rowMeans(odontocete_distance_matrix)

odontocete_sf_utm  <-  odontocete_sf_utm %>% 
  bind_cols(odo_mean_distance = odo_mean_distance)
```

```{r Mapping the mean distance}
mapview(odontocete_sf_utm, 
        zcol = "odo_mean_distance",
        layer.name = "Distance (m) from Mean Stranding Point")
```

Knowing where the "mean" stranding location is could allow for a study of that area in order to determine why odontocete strandings are so prominent on this area and allow for the allocation of more recovery resources and personnel to that area to compensate for the higher levels of strandings. Much of the strandings occurred from Charlestown to Newport, in the Rhode Island Sound.




```{r Distance Table}
# points within 40,000m
odontocete_40 <- odontocete_sf_utm %>%
  filter(odo_mean_distance < 40000) %>%
  mutate(distance_group = "< 40,000")

# points between 40,000m - 60,000m
odontocete_40_60 <- odontocete_sf_utm %>%
  filter(odo_mean_distance >= 40000 & odo_mean_distance < 60000) %>%
  mutate(distance_group = "40,000 - 60,000")

# points between 60,000m - 80,000m
odontocete_60_80 <- odontocete_sf_utm %>%
  filter(odo_mean_distance >= 60000 & odo_mean_distance < 80000) %>%
  mutate(distance_group = "60,000 - 80,000")

# points between 80,000m - 100,000m
odontocete_80_100 <- odontocete_sf_utm %>%
  filter(odo_mean_distance >= 80000 & odo_mean_distance < 100000) %>%
  mutate(distance_group = "80,000 - 100,000")

# points between 100,000m - 120,000m
odontocete_100_120 <- odontocete_sf_utm %>%
  filter(odo_mean_distance >= 100000 & odo_mean_distance < 120000) %>%
  mutate(distance_group = "100,000 - 120,000")

# points between 120,000m - 140,000m
odontocete_120_140 <- odontocete_sf_utm %>%
  filter(odo_mean_distance >= 120000 & odo_mean_distance < 140000) %>%
  mutate(distance_group = "120,000 - 140,000")

# points between 140,000m - 160,000m
odontocete_140_160 <- odontocete_sf_utm %>%
  filter(odo_mean_distance >= 140000 & odo_mean_distance < 160000) %>%
  mutate(distance_group = "140,000 - 160,000")

# points between than 160,000m - 180,000m
odontocete_160_180 <- odontocete_sf_utm %>%
  filter(odo_mean_distance >= 160000 & odo_mean_distance < 180000) %>%
  mutate(distance_group = "160,000 - 180,000")

# points greater than 180,000m
odontocete_180 <- odontocete_sf_utm %>%
  filter(odo_mean_distance >= 180000) %>%
  mutate(distance_group = "> 180,000")

# row bind the distance dataframes
odontocete_sf_utm_final <- rbind(odontocete_40, odontocete_40_60, odontocete_60_80, odontocete_80_100, odontocete_100_120, odontocete_120_140, odontocete_140_160, odontocete_160_180, odontocete_180)

odontocete_distances <- odontocete_sf_utm_final %>%
  group_by(distance_group) %>%
  summarise(totals_by_distance = n()) %>%
  st_drop_geometry() %>%
  arrange(totals_by_distance)

kable(odontocete_distances)
```

Distance (m) | Number of Odontocete Strandings
---------------|--------------------
less than 40,000 | 125
40,000 - 60,000 | 20
60,000 - 80,000 | 1
80,000 - 100,000 | 0
100,000 - 120,000 | 3
120,000 - 140,000 | 2
140,000 - 160,000 | 1
160,000 - 180,000 | 2
grater than 180,000 | 1


Of the 256 sea turtle strandings used in our analysis, 113 of them occured within 20,000 - 40,000m of the mean stranding location, followed by 92 strandings within 20,000m of the mean location (Table 1). These points were in the areas around Charlestown, RI, Newport, RI and New Shoreham, RI. This high concentration of strandings can indicate that these areas may be sea turtle stranding hotspots. This could be due to various factors, such as the flow of oceanic currents, anthropogenic impacts (such as higher fishing presrure), or the fact that there may be more manpower employed in these areas that could make stranding detection easier. Further research is needed to determine how these underlying factors may be contibuting to sea turtle strandings in these areas. 


#Temporal Analysis
Null Hypothesis: The yearly and/or monthly/seasonal odontocete stranding data is stationary
Alternative Hypothesis: The yearly and/or monthly/seasonal odontocete stranding data is not stationary

```{r Monthly Time Series}
odo_f_month <- month(first(odontocete_stranding_numbers_final$Date))
odo_f_year <- year(first(odontocete_stranding_numbers_final$Date))

odontocete_ts_monthly <- ts(odontocete_stranding_numbers_final$Total, 
                  start = c(odo_f_year, odo_f_month), 
                  frequency = 12)
 
odontocete_ts_monthly_decomp <- stl(odontocete_ts_monthly, s.window = "periodic")
plot(odontocete_ts_monthly_decomp)

odontocete_ts_monthly_trend <- Kendall::SeasonalMannKendall(odontocete_ts_monthly)

summary(odontocete_ts_monthly_trend)

```
Because the 2-sided p-value was 0.434, we fail to reject the null hypothesis and say that odontocetes monthly stranding data is stationary.

> We reject the H0 that monthly pinniped strandings are stationary, with a p-value of 1.42e-7. The decomposition indicates seasonality, which fits with the patterns that were seen in the exploratory analysis.




```{r Yearly Time Series}
odontocete_ts_yearly <- ts(odontocete_stranding_numbers_final, start = c(1990), frequency = 1)

odontocete_ts_yearly_trend <- MannKendall(odontocete_ts_yearly)

summary(odontocete_ts_yearly_trend)
```
Because the 2-sided p-value was less 7.53e-12, we reject the null hypothesis and say that odontocete yearly stranding data is not stationary. 


> We fail reject the H0 that yearly pinniped strandings are stationary with a p-value of 0.72.