---
title: "Sea Turtle Stranding Analysis"
author: "Jessica Ozog"
date: "4/2/2022"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Turtle Stranding : Set Up

```{r turtle set up}
# load packages
library(tidyverse)
library(tidyr)
library(lubridate)
library(dplyr)
library(cowplot)
library(ggplot2)
library(ggridges)
library(agricolae)
library(lubridate)
library(corrplot)
library(htmltools)
library(sf)
library(leaflet)
library(mapview)
library(RColorBrewer)
library(leafsync)
library(trend)
library(zoo)
library(Kendall)
library(tseries)
library(knitr)

# working directory and theme
getwd()
mytheme <- theme_bw() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(mytheme)

# read in the strandings data
Strandings <- read.csv("../Data/Processed/strandings_final.csv", stringsAsFactors = TRUE)
Turtle_Stranding_Data <- Strandings %>%
  filter(Family == "Turtles") %>%
  select(Family:longitude)
Turtle_Stranding_Data$Date <- as.Date(Turtle_Stranding_Data$Date, format = "%Y-%m-%d")

# strandings by YEAR (for statistical analysis)
Strandings_Totals_Year <- read.csv("../Data/Processed/strandings_totals.csv", stringsAsFactors = TRUE)
Turtle_Strandings_Year <- Strandings_Totals_Year %>%
  select(Year, Turtles_Total)

# strandings by MONTH (for statistical analysis)
Strandings_Totals_Month <- read.csv("../Data/Processed/monthly_strandings_totals.csv", stringsAsFactors = TRUE)
Turtle_Strandings_Month <- Strandings_Totals_Month %>%
  select(Month, Turtles_Total)

# turtle strandings summed by date (for time series)
Turtle_Stranding_by_date <- Turtle_Stranding_Data %>%
                    group_by(Date) %>%
                    summarise(Total = n())
```

## Turtle Stranding: Exploratory Analysis

```{r turtle exploratory analysis}
summary(Turtle_Stranding_Data)

# plot of the turtle strandings by year
ggplot(Turtle_Stranding_Data) +
  geom_bar(aes(x = Year, fill = common_name)) +
  ylab("Count") +
  labs(title = "Total Sea Turtle Strandings per Year by Species") +
  scale_fill_discrete(name = "Sea Turtle Species", labels = c("Green (Chelonia mydas)", "Kemp's Ridley (Lepidochelys kempii)", "Leatherback (Dermochelys coriacea)", "Loggerhead (Caretta caretta)", "Unidentified Turtle"))

# plot of turtle strandings by month
ggplot(Turtle_Stranding_Data) +
  geom_bar(aes(x = Month, fill = common_name)) +
  ylab("Count") +
  labs(title = "Total Sea Turtle Strandings per Month by Species") + 
  scale_fill_discrete(name = "Sea Turtle Species", labels = c("Green (Chelonia mydas)", "Kemp's Ridley (Lepidochelys kempii)", "Leatherback (Dermochelys coriacea)", "Loggerhead (Caretta caretta)", "Unidentified Turtle")) +
  scale_x_discrete(name = "Month",
                   limits = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"))
```

> Exploratory analysis of the sea turtle stranding data set by year shows higher amounts of strandings in the early 1990s and late 2000s, with the highest amount in 1995. The most prominent species throughout our study period appear to be leatherback sea turtles, followed by loggerhead sea turtles. Through monthly data exploration, strandings peak from July to September and occur less frequently the remainder of the year. 


## Turtle Stranding: Spaital Analysis

```{r turtle spatial analysis}
# create spatial frame of turtle data
Turtle_sf <- Turtle_Stranding_Data %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
mapview(Turtle_sf)

# used UTM Zone 19 (data spans across UTM zone 18 and 19)
Turtle_sf_utm <- st_transform(Turtle_sf, 32619)
mapview(Turtle_sf_utm)

# find average lat and longs
avgLong_Turtle <- mean(Turtle_Stranding_Data$longitude)
avgLat_Turtle <- mean(Turtle_Stranding_Data$latitude)
avg_Turtle_coords <- data.frame(avgLong_Turtle, avgLat_Turtle)

# center point of the turtle strandings 
avg_Turtle_sf <- avg_Turtle_coords %>%
  st_as_sf(coords = c("avgLong_Turtle", "avgLat_Turtle"), crs = 4326)
avg_Turtle_sf_utm <- st_transform(avg_Turtle_sf, 32619)
mapview(avg_Turtle_sf_utm)

# find how far each stranding is from the mean stranding location
distance_matrix <- st_distance(Turtle_sf_utm, avg_Turtle_sf_utm, pairwise=T)
#Compute the mean of each row (each EPA site)
mean_distances <-  rowMeans(distance_matrix)
#Add as a column to the epa sites
Turtle_sf_utm  <-  Turtle_sf_utm %>% 
  bind_cols(mean_distance = mean_distances) 

# make a pretty map 
mapview(Turtle_sf_utm, zcol="mean_distance", layer.name = "Distance (m) from Mean Stranding Point")
```

## Turtle Stranding: Spaital Table

```{r turtle table}
# points within 20,000m
Turtle_20 <- Turtle_sf_utm %>%
  filter(mean_distance < 20000) %>%
  mutate(distance_group = "< 20,000")

# points between 20,000m - 40,000m
Turtle_20_40 <- Turtle_sf_utm %>%
  filter(mean_distance >= 20000 & mean_distance < 40000) %>%
  mutate(distance_group = "20,000 - 40,000")

# points between 40,000m - 60,000m
Turtle_40_60 <- Turtle_sf_utm %>%
  filter(mean_distance >= 40000 & mean_distance < 60000) %>%
  mutate(distance_group = "40,000 - 60,000")

# points between 60,000m - 80,000m
Turtle_60_80 <- Turtle_sf_utm %>%
  filter(mean_distance >= 60000 & mean_distance < 80000) %>%
  mutate(distance_group = "60,000 - 80,000")

# points between 80,000m - 100,000m
Turtle_80_100 <- Turtle_sf_utm %>%
  filter(mean_distance >= 80000 & mean_distance < 100000) %>%
  mutate(distance_group = "80,000 - 100,000")

# points between 100,000m - 120,000m
Turtle_100_120 <- Turtle_sf_utm %>%
  filter(mean_distance >= 100000 & mean_distance < 120000) %>%
  mutate(distance_group = "100,000 - 120,000")

# points between 120,000m - 140,000m
Turtle_120_140 <- Turtle_sf_utm %>%
  filter(mean_distance >= 120000 & mean_distance < 140000) %>%
  mutate(distance_group = "120,000 - 140,000")

# points between 140,000m - 160,000m
Turtle_140_160 <- Turtle_sf_utm %>%
  filter(mean_distance >= 140000 & mean_distance < 160000) %>%
  mutate(distance_group = "140,000 - 160,000")

# points greater than 160,000m
Turtle_160 <- Turtle_sf_utm %>%
  filter(mean_distance >= 160000) %>%
  mutate(distance_group = "> 160,000")

# row bind the distance dataframes
Turtle_sf_utm_final <- rbind(Turtle_20, Turtle_20_40,Turtle_40_60,
                             Turtle_60_80,Turtle_80_100,Turtle_100_120,
                             Turtle_120_140,Turtle_140_160,Turtle_160)

Turtle_distances <- Turtle_sf_utm_final %>%
  group_by(distance_group) %>%
  summarise(totals_by_distance = n()) %>%
  st_drop_geometry() %>%
  arrange(totals_by_distance)

kable(Turtle_distances)
```
Distance (m) | Number of Strandings
---------------|--------------------
less than 20,000 | 92
20,000 - 40,000 | 113
40,000 - 60,000 | 32
60,000 - 80,000 | 2
80,000 - 100,000 | 4
100,000 - 120,000 | 2
120,000 - 140,000 | 6
140,000 - 160,000 | 4
greater than 160,000 | 1


> Of the 256 sea turtle strandings used in our analysis, 113 of them occured within 20,000 - 40,000m of the mean stranding location, followed by 92 strandings within 20,000m of the mean location (Table 1). These points were in the areas around Charlestown, RI, Newport, RI and New Shoreham, RI. This high concentration of strandings can indicate that these areas may be sea turtle stranding hotspots. This could be due to various factors, such as the flow of oceanic currents, anthropogenic impacts (such as higher fishing presrure), or the fact that there may be more manpower employed in these areas that could make stranding detection easier. Further research is needed to determine how these underlying factors may be contibuting to sea turtle strandings in these areas. 


## Turtle Stranding: Statistical Analysis

> Null Hypothesis 1: There is no effect of individual year on sea turtle strandings. 
> Alternative Hypothesis 1: There is an effect of individual year on sea turtle strandings. 

> Null Hypothesis 2: There is no effect of month on sea turtle strandings. 
> Alternative Hypothesis 2: There is an effect of month on sea turtle strandings. 

```{r turtles statistical analysis}
Turtle_Model_Year <- glm(data = Turtle_Strandings_Year, Turtles_Total ~ factor(Year), family = poisson)
summary(Turtle_Model_Year)

Turtle_Model_Month <- glm(data = Turtle_Strandings_Month, Turtles_Total ~ factor(Month), family = poisson)
summary(Turtle_Model_Month)
```
> We reject the null hypothesis that individual year has no effect on sea turtle strandings. The sea turtle strandings in 9 of the 21 years are not significantly different (p-values: 0.147 - 0.706). 1995 was the most significantly different (p-value: 0.000236). 1991 (p-value: 0.0056), 1993 (p-value: 0.0078), 2008 (p-value: 0.0022), 2010 (p-value: 0.0016), and 2011 (p-value: 0.003) were the next most statitistically different years. The null deviance of this model is 7.4606e+01, which indicates ??? The residual deviance of this model is -4.8850e-15, which indicates ??? 

> We reject the null hypothesis that month has no effect on sea turtle strandings. Of the 12 months, 5 of months were significantly different. May (p-value: 0.000269), June (p-value: 3.98e-06), July (p-value: 1.91e-05), August (p-value: 0.004106), and September (p-value: 0.049935) were significantly different. This indicates that sea turtle strandings are concentrated to the summer months in our study area. The null deviance of this model is 4.9235e+02, which indicates ??? The residual deviance of this model is 3.0332e-10, which indicates ???


## Turtle Stranding: Time Series

Null hypothesis:
Alternative hypothesis:

```{r turtle time series}
# time series by year
Turtle_ts_yearly <- ts(Turtle_Stranding_by_date, start = c(1990,1), frequency = 1)

# time series by month
f_month_turtle <- month(first(Turtle_Stranding_by_date$Date))
f_year_turtle <- year(first(Turtle_Stranding_by_date$Date))
Turtle_ts_monthly <- ts(Turtle_Stranding_by_date$Total, 
                  start = c(f_year_turtle, f_month_turtle), 
                  frequency = 12)
# monthly decomposition 
Turtle_ts_monthly_decomp <- stl(Turtle_ts_monthly, s.window = "periodic")
plot(Turtle_ts_monthly_decomp)

# Trends
Turtle_ts_yearly_trend <- Kendall::SeasonalMannKendall(Turtle_ts_yearly)
summary(Turtle_ts_yearly_trend)

Turtle_ts_monthly_trend <- Kendall::SeasonalMannKendall(Turtle_ts_monthly)
summary(Turtle_ts_monthly_trend)
```

```
