---
title: "Pinniped Analysis"
author: "Joshua Meza-Fidalgo"
date: '2022-04-02'
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: inline
---

### Pinniped : Set Up

```{r Initial setup, warning = FALSE, message = FALSE}
library(tidyverse)
library(tidyr)
library(lubridate)
library(dplyr)
library(cowplot)
library(ggplot2)
library(ggridges)
library(agricolae)
library(lubridate)
library(corrplot)
library(htmltools)
library(sf)
library(leaflet)
library(mapview)
library(RColorBrewer)
library(leafsync)
library(trend)
library(zoo)
library(Kendall)
library(tseries)
library(knitr)

getwd()

mytheme <- theme_bw() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(mytheme)
```



```{r Data Setup}
#Reading in initial dataset
Stranding_Data <- read.csv("../Data/Processed/strandings_final.csv", stringsAsFactors = TRUE)

#Wrangling for pinniped only data
Pinniped_Stranding_Data <- Stranding_Data %>%
  filter(Family == "Pinnipeds") %>%
  select(Family:longitude)


Pinniped_Stranding_Data$Date <- as.Date(Pinniped_Stranding_Data$Date, format = "%Y-%m-%d")


Pinniped_Stranding_Totals <- Pinniped_Stranding_Data %>%
                    filter(Family == "Pinnipeds") %>%
                    group_by(Month, Year) %>%
                    summarise(Total = n())


#Reading in and wrangling strandings by year
Annual_Strandings_Totals <- read.csv("../Data/Processed/annual_strandings_totals.csv", stringsAsFactors = TRUE)

Annual_Pinniped_Strandings <- Annual_Strandings_Totals %>%
  select(Year, Pinnipeds_Total)



#Reading in and wrangling strandings by month
Monthly_Strandings_Totals <- read.csv("../Data/Processed/monthly_strandings_totals.csv", stringsAsFactors = TRUE)

Monthly_Pinniped_Strandings <- Monthly_Strandings_Totals %>%
  select(Month, Pinnipeds_Total)
```


## Pinniped Stranding Exploratory Analysis

```{r Pinniped Data Exploration}
summary(Pinniped_Stranding_Data)

#Plot of annual pinniped strandings
ggplot(Pinniped_Stranding_Data) +
  geom_bar(aes(x = Year, fill = common_name)) +
  ylab("Count") +
  labs(title = "Total Pinniped Strandings per Year by Species") + 
  scale_fill_discrete(name = "Pinniped Species", labels = c("Atlantic Gray Seal", "Harbor Seal", "Harp Seal", "Hooded Seal", "Unidentified Pinniped"))

#Plot of overall monthly pinniped strandings
ggplot(Pinniped_Stranding_Data) +
  geom_bar(aes(x = Month, fill = common_name)) +
  ylab("Count") +
  labs(title = "Total Pinniped Strandings per Month by Species") + 
  scale_fill_discrete(name = "Pinniped Species", labels = c("Atlantic Gray Seal", "Harbor Seal", "Harp Seal", "Hooded Seal", "Unidentified Pinniped")) +
  scale_x_discrete(name = "Month",
                   limits = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"))
```


```{r, message = FALSE}
#Plot of annual pinniped strandings with trend line
Pinniped_Stranding_Plot <-
ggplot(Annual_Pinniped_Strandings, aes(x = Year, y = Pinnipeds_Total)) +
  geom_line(color = "red") +
  ylab("Pinniped Strandings") +
  geom_smooth(method = lm) +
  labs(title = "Pinniped Strandings Per Year")

print(Pinniped_Stranding_Plot)
```

>Pinniped data exploration indicates a possible overall increasing trend in pinniped strandings and a possible monthly/seasonal trend.  There is some fluctuation in number of total yearly strandings both for the family as a whole and at the species level, although a more detailed species level analysis is beyond the scope of this project. Harp and harbor seals make up most of the strandings each year. Most yearly strandings occur in the late winter and spring time, particularly in March, with very few in the late summer through the early winter.  Springtime coincides with the pupping seasons for many seal species and could be a factor in this.



```{r Geospatial Exploration}
#Spatial frame to visualize the stranding data spatially.
Pinniped_sf <- Pinniped_Stranding_Data %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

mapview(Pinniped_sf)
```
Pinniped strandings are most numerous throughout the coastline of Rhode Island and the eastern coastline of Connecticut.

### Pinniped Stranding Analysis


## Pinniped Statisical Analysis

>Null Hypothesis 1: There is no effect of year on the total number of pinniped strandings
>Alternative Hypothesis 1: There is an effect of year on the total number of pinniped strandings

>Null Hypothesis 2: There is no effect of month on the total number of pinniped strandings
>Alternative Hypothesis 2: There is an effect of month on the total number of pinniped strandings

```{r Poisson Models}
#Developing an model to analysis the significance of year on pinniped strandings
Annual_Pinniped_Model <- glm(data = Annual_Pinniped_Strandings, Pinnipeds_Total ~ factor(Year), family = poisson)

summary(Annual_Pinniped_Model)


#Developing an model to analysis the significance of month on pinniped strandings
Monthly_Pinniped_Model <- glm(data = Monthly_Pinniped_Strandings, Pinnipeds_Total ~ factor(Month), family = poisson)

summary(Monthly_Pinniped_Model)
```

> We reject the null hypothesis that there is no effect of year on the total number of pinniped strandings.  18 of the 21 documented years were statistically significant with p-values ranging from 9.07e-09(2001) to 0.03(1999).  Only the first three years of data, 1991-1993 had p-values above 0.05.  The null deviance of yearly model was 1.36e+02 on 21 degrees of freedom, and residual deviance of 1.31e-14

> We reject the null hypothesis that there is no effect of month on the total number of pinniped strandings.  11 of the 12 months had p-values of less than 0.05, with July being the exception.  The p-values of the statistically significant month ranged from 1.77e-15(March) to 0.02(June).  The null deviance for the monthly model was 5.95e+02 on 11 degrees of freedom and a residual deviance of -3.55e-15. 



## Pinniped Geospatial Analysis

```{r}
#Converting the data points to UTM for further analysis. The study are straddles UTM zones 18 and 19, zone 19 was chosen because that is where the greatest concentration of points seems to be located.

Pinniped_sf_utm <- st_transform(Pinniped_sf,32619)

Pinniped_sf_utm$geometry

#Determining "mean" stranding point for distance comparisons
long_mean <- mean(Pinniped_Stranding_Data$longitude)
lat_mean <- mean(Pinniped_Stranding_Data$latitude)

Pinniped_Stranding_Mean_Point <- data.frame(long_mean, lat_mean)

Pinniped_Mean_sf <- Pinniped_Stranding_Mean_Point %>%
  st_as_sf(coords = c("long_mean", "lat_mean"), crs = 4326)

Pinniped_Mean_UTM <- st_transform(Pinniped_Mean_sf,32619)

#Creating a distance matrix for further analysis
Pinniped_Distance_Matrix <- st_distance(Pinniped_sf_utm, Pinniped_sf_utm, pairwise = T)

mean_distance <-  rowMeans(Pinniped_Distance_Matrix)

Pinniped_sf_utm  <-  Pinniped_sf_utm %>% 
  bind_cols(mean_distance = mean_distance)
```

```{r}
#Visualizing stranding point distance from "mean"
mapview(Pinniped_sf_utm, 
        zcol = "mean_distance",
        layer.name = "Distance (m) from Mean Stranding Point")
```


```{r Distance Table}
#Developing a mean distance table

# points within 40,000m
Pinniped_40 <- Pinniped_sf_utm %>%
  filter(mean_distance < 40000) %>%
  mutate(distance_group = "< 40,000")

# points between 40,000m - 60,000m
Pinniped_40_60 <- Pinniped_sf_utm %>%
  filter(mean_distance >= 40000 & mean_distance < 60000) %>%
  mutate(distance_group = "40,000 - 60,000")

# points between 60,000m - 80,000m
Pinniped_60_80 <- Pinniped_sf_utm %>%
  filter(mean_distance >= 60000 & mean_distance < 80000) %>%
  mutate(distance_group = "60,000 - 80,000")

# points between 80,000m - 100,000m
Pinniped_80_100 <- Pinniped_sf_utm %>%
  filter(mean_distance >= 80000 & mean_distance < 100000) %>%
  mutate(distance_group = "80,000 - 100,000")

# points between 100,000m - 120,000m
Pinniped_100_120 <- Pinniped_sf_utm %>%
  filter(mean_distance >= 100000 & mean_distance < 120000) %>%
  mutate(distance_group = "100,000 - 120,000")

# points between 120,000m - 140,000m
Pinniped_120_140 <- Pinniped_sf_utm %>%
  filter(mean_distance >= 120000 & mean_distance < 140000) %>%
  mutate(distance_group = "120,000 - 140,000")

# points between 140,000m - 160,000m
Pinniped_140_160 <- Pinniped_sf_utm %>%
  filter(mean_distance >= 140000 & mean_distance < 160000) %>%
  mutate(distance_group = "140,000 - 160,000")

# points greater than 160,000m
Pinniped_160 <- Pinniped_sf_utm %>%
  filter(mean_distance >= 160000) %>%
  mutate(distance_group = "> 160,000")

# row bind the distance dataframes
Pinniped_sf_utm_final <- rbind(Pinniped_40, Pinniped_40_60, Pinniped_60_80, Pinniped_80_100, Pinniped_100_120, Pinniped_120_140, Pinniped_140_160, Pinniped_160)

Pinniped_distances <- Pinniped_sf_utm_final %>%
  group_by(distance_group) %>%
  summarise(totals_by_distance = n()) %>%
  st_drop_geometry() %>%
  arrange(totals_by_distance)

#kable(Pinniped_distances)
#Table Reorganized manually based on kable results
```
Distance (m) | Number of Strandings
---------------|--------------------
20,000 - 40,000 | 425
40,000 - 60,000 | 200
60,000 - 80,000 | 15
80,000 - 100,000 | 11
100,000 - 120,000 | 7
120,000 - 140,000 | 11
140,000 - 160,000 | 7
greater than 160,000 | 3


> By finding the "mean" stranding location based on the all of the data points it can be determined how far each stranding point is from the mean.  By determining that 425 out of 679, or 62.5 percent of, pinniped strandings occur within 40km of the mean stranding location, what makes this area a hotspot can be studied and hopefully also determined.  This data also could allow for reallocation of stranding resources and personnel to this range in order to put them to the most efficient use possible.  Since only 54, or 8 percent of, pinniped strandings occur further than 60km from the stranding mean, resources for study and rescue can be allocated away from those areas.


## Pinniped Temporal Analysis

>Null Hypothesis: The yearly and/or monthly/seasonal pinniped stranding data is stationary
>Alternative Hypothesis: The yearly and/or monthly/seasonal pinniped stranding data is not stationary

```{r Monthly Time Series}
#Developing a monthly time series to determine if there is a monthly/seasonal trend.
f_month <- month(first(Pinniped_Stranding_Totals$Month))


Pinniped_ts_monthly <- ts(Pinniped_Stranding_Totals$Total, 
                  start = c(1990, f_month), 
                  frequency = 12)
 
Pinniped_ts_monthly_decomp <- stl(Pinniped_ts_monthly, s.window = "periodic")
plot(Pinniped_ts_monthly_decomp)

Pinniped_ts_monthly_trend <- Kendall::SeasonalMannKendall(Pinniped_ts_monthly)

summary(Pinniped_ts_monthly_trend)
```

> We reject the H0 that monthly pinniped strandings are stationary, with a p-value of 1.42e-7. The decomposition indicates seasonality, which fits with the patterns that were seen in the exploratory analysis.



```{r Yearly Time Series}
#Developing a yearly time series to determine if there is a trend
Pinniped_ts_yearly <- ts(Pinniped_Stranding_Totals, start = c(1990), frequency = 1)

Pinniped_ts_yearly_trend <- MannKendall(Pinniped_ts_yearly)

summary(Pinniped_ts_yearly_trend)
```

> We fail reject the H0 that yearly pinniped strandings are stationary with a p-value of 0.72.





Pinniped Summary:

While there were hints of a possible increasing trend in yearly pinniped strandings, a time series analysis found this to not be the case.  Additionally, 85.7 percent of the years were statistically significant for pinniped strandings.  There was seasonality in pinniped strandings, along with all but one of the months being statistically significant, which could be attributed to pupping seasons.  62.5 percent of pinniped strandings happened within 40km of the "mean stranding point."  By knowing where and when pinnipeds are most likely to strand, the environmental factors of those places and times can be studied in order to determine the conditions that cause strandings.  This knowledge could aid in stranding prevention and/or mitigation measures.  Further study could also be conducted at the species level to determine what if any differences occur.
